---
title: "Hierarchical Bayesian Estimation of the Circular Diffusion Model with EMC2"
author: "Amir Hosein Hadian"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 5
)
```

# Overview

This tutorial demonstrates how to estimate a **hierarchical Bayesian** Circular Diffusion Model (CDM; Smith, 2016) using the `EMC2` package.

The model is fitted simultaneously to data from multiple participants. In a hierarchical Bayesian model, participant-level parameters are estimated together with group-level distributions. This allows information to be shared across participants while still preserving individual differences.

# 1. Load packages and set the random seed

We begin by clearing the workspace, loading `EMC2`, and setting a random seed so that stochastic parts of the analysis are reproducible.

```{r load-packages}
rm(list = ls())

library(EMC2)

set.seed(123456)
```

# 2. Load the data

The data are stored in `Kvam2019.csv`.

```{r load-data}
data <- read.csv(
  "_data/Kvam2019.csv",
  header = TRUE,
  stringsAsFactors = FALSE
)
```

Inspect the first few rows:

```{r inspect-data}
head(data)
```

The hierarchical analysis requires a participant identifier. Here, the `subjects` column identifies individual participants.

# 3. Prepare categorical variables

We convert the relevant variables to factors.

```{r prepare-factors}
data$subjects <- factor(data$subjects)

data$isSpeed <- factor(
  data$isSpeed,
  levels = c(0, 1),
  labels = c("accuracy", "speed")
)

data$jitter <- factor(data$jitter)
```

It is useful to verify the resulting factor levels before fitting the model.

```{r inspect-factors}
levels(data$subjects)
levels(data$isSpeed)
levels(data$jitter)
```


# 4. Visualize the observed data

Before fitting a cognitive model, it is useful to inspect the empirical response distributions.

```{r accuracy-function}
acc_fun <- function(data) {
  return(data$abs_error <= pi)
}
```

We can then plot the observed response-time density separately for the different jitter conditions.

```{r observed-density, fig.width=8, fig.height=5}
plot_density(
  data,
  factors = c("isSpeed", "jitter"),
  functions = c(correct = acc_fun),
  defective_factor = "correct",
  layout = c(2, 3)
)
```

```{r observed-density, fig.width=8, fig.height=5}
plot_density(
  data,
  density_over = "R",
  factors = c("isSpeed", "jitter"),
  layout = c(2, 3)
)
```

These plots provide an initial description of how response times and response errors vary across stimulus conditions.

# 5. Specify the circular diffusion model

The experimental manipulations are incorporated into the CDM through the model design.

```{r model-design}
design_CDM_SAT <- design(
  data = data,
  Rlevels = "R",
  formula = list(
    v     ~ 0 + jitter,
    theta ~ 1,
    a     ~ isSpeed,
    t0    ~ 1,
    sv    ~ 1
  ),
  model = CDM 
)
```

The formulas determine which parameters vary across experimental conditions.

- `v ~ 0 + jitter` estimates a separate drift magnitude for each jitter (difficulty) level. Removing the intercept gives the parameters `v_jitter15`, `v_jitter30`, and `v_jitter45` directly.
- `theta ~ 1` estimates one common drift angle across all conditions.
- `a ~ isSpeed` estimates a baseline threshold for the reference level (`accuracy`) plus a speed-condition effect. Thus, the model allows response caution to differ between speed and accuracy instructions.
- `t0 ~ 1` estimates one common non-decision time.
- `sv ~ 1` estimates an across-variability in drift rate.

# 6. Inspect the model design

Before fitting the model, it is important to confirm that the design matrix and parameter mapping are what we intended.

```{r inspect-design}
print(design_CDM_SAT)
```

A more detailed summary can be obtained with:

```{r design-summary}
summary(design_CDM_SAT)
```

The parameters that will be sampled can be inspected using:

```{r sampled-parameters}
sampled_pars(design_CDM_SAT)
```

This step is especially useful when a model contains several factor-dependent parameters.

# 7. Hierarchical Bayesian estimation

The EMC object is created using the complete data set.

```{r make-emc}
emc <- make_emc(
  data = data,
  design = design_CDM_SAT
)
```

Because all participants are included in `data`, the model is fitted hierarchically rather than separately for each participant.

The hierarchical structure provides partial pooling: participant-level estimates are informed by both the participant's own data and the population-level distribution.

# 8. Fit the hierarchical model

We now fit the model.

```{r fit-model}
fit_model <- FALSE

if(fit_model){
  hier_CDM_emc <- fit(emc)
  save(hier_CDM_emc, file="_samples/hier_CDM_emc.RData")
} else{
  print(load("_samples/hier_CDM_emc.RData"))
}



```

Depending on the number of participants, trials, and MCMC settings, hierarchical estimation can be computationally intensive.

# 9. Check sampling and convergence

After fitting the model, the first step is to inspect the sampler diagnostics.

```{r check-model}
check(hier_CDM_emc)
```

The purpose of this step is to determine whether the posterior sampler has adequately explored the posterior distribution.


# 10. Plot the chains

Posterior distributions can be visualized directly.

```{r posterior-plots, fig.width=9, fig.height=6}
plot(hier_CDM_emc)
```

These plots are useful for examining posterior uncertainty and identifying possible sampling problems.

# 11. Credible intervals

Posterior credible intervals can be obtained using:

```{r credible-intervals}
credint(
  hier_CDM_emc,
  map = TRUE
)
```

# 12. Group- and participant-level parameter estimates

One of the main advantages of hierarchical estimation is that the model provides both population-level and participant-level parameter estimates.

The following plot displays the hierarchical parameter estimates while also showing individual participants:

```{r subject-parameters, fig.width=10, fig.height=7}
plot_pars(
  hier_CDM_emc,
  map = TRUE,
  all_subjects = TRUE,
  use_prior_lim = FALSE
)
```

This visualization makes it possible to examine individual differences relative to the population-level posterior.

Participant estimates in a hierarchical model are partially pooled toward the population distribution. Participants with less informative data generally receive more shrinkage than participants whose data strongly identify their parameters.

# 13. Posterior predictive simulation

Posterior predictive checks evaluate whether data generated from the fitted model resemble the observed data.

We first generate posterior predictive data.

```{r posterior-prediction}
sim_model <- FALSE
if (sim_model){
  pp_hier_CDM <- predict(
    hier_CDM_emc,
    n_cores = 4,
    n_post = 50
  )
  save(pp_hier_CDM, file="pp_hier_CDM.RData")
} else{
  print(load("pp_hier_CDM.RData"))
}

```


Here:

- `n_cores = 4` uses four CPU cores for posterior prediction;
- `n_post = 50` draws predictions from fifty posterior samples.

For a final analysis, a larger value of `n_post` may provide a more stable representation of posterior predictive uncertainty.

Inspect the structure of the simulated data:

```{r inspect-posterior-prediction}
head(pp_hier_CDM)
```

# 14. Posterior predictive density check

We can compare observed and posterior-predictive response-time distributions across both jitter and speed--accuracy conditions.

```{r posterior-density, fig.width=10, fig.height=8}
plot_density(
  hier_CDM_emc,
  pp_hier_CDM,
  functions = c(correct = acc_fun),
  defective_factor = "correct",
  factors = c("isSpeed", "jitter"),
  layout = c(2, 3)
)
```

```{r posterior-density, fig.width=10, fig.height=8}
plot_density(
  hier_CDM_emc,
  pp_hier_CDM,
  density_over = "R",
  factors = c("isSpeed", "jitter"),
  layout = c(2, 3)
)
```

A well-fitting model should reproduce the major characteristics of the empirical response-time and response angle distributions across conditions.

# 15. Posterior predictive CDF check

The cumulative distribution function provides another useful posterior predictive diagnostic.

```{r posterior-cdf, fig.width=10, fig.height=8}
plot_cdf(
  hier_CDM_emc,
  pp_hier_CDM,
  factors = c("isSpeed", "jitter"),
  functions = c(correct = acc_fun),
  defective_factor = "correct",
  layout = c(2, 3)
)
```

CDF plots are particularly useful for identifying systematic discrepancies across different portions of the response-time distribution.

# References

Kvam, P. D. (2019). Modeling accuracy, response time, and bias in continuous orientation judgments. *Journal of Experimental Psychology: Human Perception and Performance*, **45**(3), 301--318. <https://doi.org/10.1037/xhp0000606>

Smith, P. L. (2016). Diffusion theory of decision making in continuous report. *Psychological Review*, **123**(4), 425–451. <https://doi.org/10.1037/rev0000023>
