---
title: "Bayesian Estimation of the Circular Diffusion Model with EMC2"
author: "Amir Hosein Hadian"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 5
)
```

# Overview

This tutorial presents a **Bayesian** estimation of the Circular Diffusion Model
(CDM; Smith, 2016) using the `EMC2` to model the continuous orientation-judgment data from Kvam (2019) and demonstrates how experimental manipulations can be mapped onto condition-specific model parameters.

The experiment contains two manipulations:

- **Speed--accuracy instruction** (`isSpeed`): accuracy versus speed blocks.
- **Stimulus difficulty** (`jitter`): orientation jitter of 15, 30, or 45 degrees.

To estimate the model we assume a CDM with:

- drift magnitude, $v$, varies across the three jitter levels;
- decision threshold, $a$, varies between speed and accuracy conditions;
- drift angle, $\theta$, is shared across conditions;
- non-decision time, $t_0$, is shared across conditions;

# 1. Load EMC2 and other required packages

```{r packages}
library(EMC2)

set.seed(123456)
```

> **Important:** This tutorial assumes an `EMC2` installation/build in which the Circular Diffusion Model is available as `CDM`, as in the accompanying R script. If `CDM` is not found in your installation, please use the development version.

# 2. Load and prepare the data

The Kvam (2019) data contains the following columns:

- `subjects`: participant identifier;
- `isSpeed`: instruction condition, coded `0 = accuracy` and `1 = speed` in the raw file;
- `jitter`: stimulus jitter in degrees (15, 30, or 45);
- `abs_error`: absolute angular response error in radians;
- `R`: signed angular response error in radians;
- `rt`: response time in seconds.


```{r load-data}
possible_paths <- c(
  "_data/Kvam2019.csv"
)

data_path <- possible_paths[file.exists(possible_paths)][1]

if (is.na(data_path)) {
  stop(
    "Could not find the Kvam2019 data file. Place it in '_data/Kvam2019.csv' ",
    "or in the same directory as this R Markdown file."
  )
}

data <- read.csv(
  data_path,
  header = TRUE,
  stringsAsFactors = FALSE
)

# Convert experimental variables to factors
data$subjects <- factor(data$subjects)
data$isSpeed <- factor(
  data$isSpeed,
  levels = c(0, 1),
  labels = c("accuracy", "speed")
)
data$jitter <- factor(data$jitter)

head(data)
```

A quick summary of the design is useful before fitting the model.

```{r data-summary}
cat("Number of participants:", nlevels(data$subjects), "\n")
cat("Number of trials:", nrow(data), "\n\n")

with(data, table(jitter, isSpeed))
```

# 3. Select one participant

The current example is a **single-level Bayesian analysis**. We therefore begin with the a single participant.

```{r select-participant}
single_subject <- levels(data$subjects)[3]

data_single <- droplevels(
  data[data$subjects == single_subject, , drop = FALSE]
)

cat("Participant:", single_subject, "\n")
cat("Number of trials:", nrow(data_single), "\n")
```

To visualize the behavioral data, we define a response as relatively accurate when its absolute angular error is no larger than 10 degrees ($\pi/18$ radians). This is used only for plotting and posterior-predictive summaries; it does **not** convert the continuous-response model into a binary-choice model.

```{r observed-density, fig.height=5}
acc_fun <- function(data) {
  data$abs_error <= pi
}

plot_density(
  data_single,
  factors = "jitter",
  functions = c(correct = acc_fun),
  defective_factor = "correct"
)
```

# 4. Specify the Circular Diffusion Model

The key step is to translate the experimental design into linear-model formulas for the CDM parameters.

```{r check-CDM}
?CDM
```

```{r model-design}
design_CDM_SAT <- design(
  data = data_single,
  Rlevels = "R",
  formula = list(
    v     ~ 0 + jitter,
    theta ~ 1,
    a     ~ isSpeed,
    t0    ~ 1
  ),
  model = CDM
)
```

- `v ~ 0 + jitter` estimates a separate drift magnitude for each jitter (difficulty) level. Removing the intercept gives the parameters `v_jitter15`, `v_jitter30`, and `v_jitter45` directly.
- `theta ~ 1` estimates one common drift angle across all conditions.
- `a ~ isSpeed` estimates a baseline threshold for the reference level (`accuracy`) plus a speed-condition effect. Thus, the model allows response caution to differ between speed and accuracy instructions.
- `t0 ~ 1` estimates one common non-decision time.
- The model does not include any across-variability.

We can inspect the resulting design and the parameters that EMC2 will sample.

```{r inspect-design}
print(design_CDM_SAT)
summary(design_CDM_SAT)
sampled_pars(design_CDM_SAT)
```

# 5. Specify Bayesian priors

In the optimization-based JEAM analysis, parameter ranges constrain the search space. In Bayesian estimation, we instead express plausible parameter values through prior distributions.

The prior means below reproduce the values from the accompanying EMC2 script. Positive parameters such as drift magnitude, threshold, and non-decision time are centered on the transformed scale used by the model.

```{r priors}
p_vector <- c(
  v_jitter15       = log(2),
  v_jitter30       = log(1.5),
  v_jitter45       = log(1),
  theta            = qnorm(0.5),
  a                = log(1),
  a_isSpeedspeed   = -0.2,
  t0               = log(0.3)
)

prior_sd <- c(
  v_jitter15       = 1,
  v_jitter30       = 1,
  v_jitter45       = 1,
  theta            = 0.3,
  a                = 2,
  a_isSpeedspeed   = 0.2,
  t0               = 0.4
)

prior_CDM_SAT <- prior(
  design_CDM_SAT,
  pmean = p_vector,
  psd = prior_sd,
  type = "single"
)

prior_CDM_SAT
```

These priors encode the expectation that evidence quality decreases as stimulus jitter increases: the prior center for drift magnitude is highest for jitter 15, intermediate for jitter 30, and lowest for jitter 45. The negative prior center for `a_isSpeedspeed` expresses the expectation that the decision threshold is lower under speed emphasis than under accuracy emphasis.

# 6. Create the EMC2 object and fit the model

We now combine the data, design, and priors into an EMC2 object and sample from the posterior distribution.

```{r fit-model, cache=TRUE}
fit_model <- FALSE

emc <- make_emc(
  data = data_single,
  design = design_CDM_SAT,
  prior = prior_CDM_SAT,
  type = "single"
)

if(fit_model){
  CDM_emc <- fit(emc)
  save(CDM_emc,file="_samples/CDM_emc.RData")
} else{
  print(load("_samples/CDM_emc.RData"))
}
```

Because Bayesian estimation is based on Monte Carlo sampling, convergence should be checked before interpreting the posterior.

# 7. Check sampling and convergence

```{r convergence-check}
check(CDM_emc)
```

Trace and marginal posterior plots provide a visual diagnostic of the sampler.

```{r posterior-traces, fig.height=7}
plot(CDM_emc)
```

Poor mixing, strong trends across iterations, or disagreement between chains should be addressed before interpreting parameter estimates or posterior predictions.

# 8. Summarize the posterior distributions

Posterior credible intervals quantify uncertainty in each parameter. Setting `map = TRUE` reports parameters on their mapped/interpretable scale where supported by the model.

```{r credible-intervals}
credint(CDM_emc, map = TRUE)
```

The posterior distributions can also be visualized directly.

```{r posterior-parameters, fig.height=6}
plot_pars(CDM_emc, map = TRUE, use_prior_lim = FALSE)
```

# 9. Posterior-predictive simulation

A fitted model should reproduce important properties of the observed data, not merely yield plausible parameter estimates. We therefore generate posterior-predictive datasets.

```{r posterior-prediction, cache=TRUE}
sim_model <- FALSE
if (sim_model){
  pp_CDM <- predict(
    CDM_emc,
    n_cores = 4,
    n_post = 50
  )
  save(pp_CDM, file="pp_CDM.RData")
} else{
  print(load("pp_CDM.RData"))
}

head(pp_CDM)
```

`n_post` controls the number of posterior draws used for prediction. For a publication-quality posterior-predictive assessment, it is generally useful to increase this value once the model-fitting workflow has been verified.

# 10. Posterior-predictive checks

## 10.1 Response time density:

The following plot compares observed and posterior-predicted response distributions across jitter and speed--accuracy conditions.

```{r posterior-density, fig.height=8}
plot_density(
  CDM_emc,
  pp_CDM,
  factors = c("isSpeed", "jitter"),
  functions = c(correct = acc_fun),
  defective_factor = "correct",
  layout = c(2, 3)
)
```

## 10.2 Response angle density:
```{r posterior-density, fig.height=8}
plot_density(
  CDM_emc,
  pp_CDM,
  density_over = "R",
  factors = c("isSpeed", "jitter"),
  layout = c(2, 3)
)
```


## 10.3 Response time CDF:

We can also examine whether the posterior predictions reproduce the shape and condition dependence of the response-time distributions.

```{r posterior-cdf, fig.height=8}
plot_cdf(
  CDM_emc,
  pp_CDM,
  factors = c("isSpeed", "jitter"),
  functions = c(correct = acc_fun),
  defective_factor = "correct",
  layout = c(2, 3)
)
```

Together, the response time and response angle densities and CDF checks assess whether the CDM jointly accounts for **continuous response error and response time**, which is the central advantage of the model.

# References

Kvam, P. D. (2019). Modeling accuracy, response time, and bias in continuous orientation judgments. *Journal of Experimental Psychology: Human Perception and Performance*, **45**(3), 301--318. <https://doi.org/10.1037/xhp0000606>

Smith, P. L. (2016). Diffusion theory of decision making in continuous report. *Psychological Review*, **123**(4), 425–451. <https://doi.org/10.1037/rev0000023>




