---
title: "Fitting a Hierarchical Model with Between-Subject effects in EMC2"
description: "Data preparation, model specification, estimation, and interpretation in EMC2"
output: rmarkdown::html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
rm(list = ls())
library(EMC2)
knitr::opts_chunk$set(comment = "#>", message = FALSE, warning = FALSE)
```

## Introduction

This vignette shows how to fit a hierarchical Linear Ballistic
Accumulator (LBA) model in `EMC2` to the direction integration task
presented in Manning et al. (2022).

-   an `EMC2` design object,
-   a hierarchical group-level model,
-   a prior specification,
-   a fitted `emc` object,
-   and finally a set of posterior summaries and posterior predictive
    checks.

The analysis here uses the `deviation` task and compares the `CTRL` and
`ASD` groups across `hard` and `easy` trials.

## Load and prepare the data

The first job is to put the dataset into the format that `EMC2` expects:

-   `subjects`: participant identifier,
-   `R`: the observed response,
-   `rt`: response time in seconds.

In the raw Manning2022 file these variables have different names, so we
recode them explicitly. We further recode factors in the data like `S`
(stimulus) and `group` to have interpretable levels.

```{r data-prep}
load("data/Manning2022.RData")

data <- subset(
  data,
  task == "deviation" & !is.na(RT),
  select = c(cond, motiondir, keypress, RT, subjects, group)
)

colnames(data) <- c("cond", "S", "R", "rt", "subjects", "group")

data$R <- ifelse(data$R == "2", "right", "left")
data$group <- ifelse(data$group == 0, "CTRL", "ASD")

data$subjects <- factor(data$subjects)
data$cond <- factor(data$cond, levels = c("hard", "easy"))
data$R <- factor(data$R)
data$group <- factor(data$group, levels = c("CTRL", "ASD"))
data$S <- factor(data$S)
```

Before fitting anything, it is worth checking that the dataset still
looks sensible after filtering and recoding.

```{r data-overview}
c(
  n_trials = nrow(data),
  n_subjects = nlevels(data$subjects)
)

with(data, table(group, cond))
with(data, table(group, S))
```

The trial counts are well balanced across groups and conditions.

## Inspect the observed data

It is good practice to look at the data to see if there is any
irregularities. `plot_density()` is useful here because it plots
defective densities: the area under each density is proportional to
the probability of that response class, so accuracy and response-time
shape are visualized together.

I define a small helper that classifies each trial as correct or
incorrect, then plot the distributions separately for condition and
group.

```{r observed-densities, fig.height=6, fig.alt="Observed defective densities by condition and group"}
acc_fun <- function(data) factor(data$S == data$R)

plot_density(
  data,
  factors = c("cond", "group"),
  functions = list(correct = acc_fun),
  defective_factor = "correct"
)
```

Here we use a response time cutoff between 250 ms and 2 seconds.

```{r trim-rt}
range(data$rt)

data <- subset(data, rt < 2 & rt > 0.25)
```

This is not a universal rule. It is just the trimming choice for this
analysis. In your own work, outlier handling should be justified from
the task and measurement setup rather than copied blindly.

## LBA overview

The LBA is a race model with one accumulator per response option. In
this task there are two response options, so each trial is modeled as a
race between a left and a right accumulator.

The parameters used below are:

-   `v`: mean drift rate,
-   `B`: threshold distance,
-   `A`: start-point range,
-   `t0`: non-decision time,
-   `sv`: between-trial drift-rate variability.

In `EMC2`, race models create an `lR` factor, which has the same levels
as the response factor `R`. These correspond to the accumulators present
on each trial (a left and a right accumulator). This allows us to vary
the design as a function of the accumulator. For example we can set
`B ~ lR` to allow for different levels of evidence required for the left
or right accumulator, which is a way to model response bias. We can
further use a `matchfun` to code which accumulator matches the stimulus
and should therefore be the favored response.

## Specify the subject-level design

For binary matching effects it is often useful to use an
average-difference contrast. To that end we set up the following
contrast matrix. Its effect will be illustrated below

```{r contrast-matrix}
ADmat <- cbind(d = c(-1 / 2, 1 / 2))
ADmat
```

Now we specify the subject-level model.

```{r subject-design}
des <- design(
  formula = list(v ~ lM * cond, B ~ lR, t0 ~ 1, A ~ 1, sv ~ lM),
  data = data,
  matchfun = function(data) data$S == data$lR,
  model = LBA,
  contrasts = list(lM = ADmat),
  constants = c(sv = log(1))
)

sampled_pars(des)
```

There are several important modeling choices here.

-   `v ~ lM * cond` says that drift rate depends on whether an
    accumulator matches the correct response, and that this matching
    effect may change between hard and easy trials.
-   `B ~ lR` allows different thresholds for the left and right
    accumulators, which is the LBA analogue of response bias.
-   `t0 ~ 1` and `A ~ 1` estimate non-decision time and start-point
    range as consistent across cells of the design
-   `sv ~ lM` allows a match-related difference in drift variability,
    but the intercept of `sv` is fixed with `constants = c(sv = log(1))`
    to satisfy the scaling constraint. The constant is set on the
    sampled scale (log).

The output of `sampled_pars(des)` is worth reading carefully because it
tells you exactly which coefficients the model will estimate:

-   `v` is the overall drift-rate intercept.
-   `v_lMd` is the matching-minus-mismatching drift-rate contrast.
-   `v_condeasy` is the easy-minus-hard drift-rate shift.
-   `v_lMd:condeasy` is the interaction: does the matching advantage
    change on easy trials?
-   `B_lRright` is the extra threshold for the right accumulator
    relative to the left accumulator.

The printed design matrices also show how these coefficients are
combined on each trial type. For example, the `v` design implies:

$$
v_{\mathrm{match,hard}} = v + 0.5\,v_{lMd}
$$

$$
v_{\mathrm{mismatch,hard}} = v - 0.5\,v_{lMd}
$$

$$
v_{\mathrm{match,easy}} = v + v_{condeasy} + 0.5\,v_{lMd} + 0.5\,v_{lMd:condeasy}
$$

$$
v_{\mathrm{mismatch,easy}} = v + v_{condeasy} - 0.5\,v_{lMd} - 0.5\,v_{lMd:condeasy}
$$

That is exactly the kind of bookkeeping `design()` is supposed to handle
for you.

## Add the hierarchical group model

The subject-level design tells `EMC2` which parameters exist for each
participant. `group_design()` then tells `EMC2` which of those
subject-level parameters should vary systematically with
participant-level predictors.

Here the only participant-level predictor is diagnostic group.

```{r group-design}
group_des <- group_design(
  list(
    v ~ group,
    v_lMd ~ group,
    v_condeasy ~ group,
    `v_lMd:condeasy` ~ group,
    B ~ group,
    t0 ~ group
  ),
  subject_design = des,
  data = data
)

sampled_pars(group_des)
```

`group_design()` does not remove the hierarchical intercept. It adds
predictors to the group-level means of the selected subject-level
parameters.

That means:

-   `v` is still the mean subject-level drift intercept for the
    reference group,
-   `v_groupASD` is the ASD-minus-CTRL shift in that drift intercept,
-   `t0_groupASD` is the ASD-minus-CTRL shift in non-decision time,
-   and so on.

Parameters that are **not** named in `group_design()` still get a
hierarchical intercept and variance, but they do not get an explicit
group regression term. In this model, for example, `A`, `B_lRright`, and
`sv_lMd` are not modeled as a function of group.

## Specify priors

Calling `prior()` after defining `des` and `group_des` creates priors
for the hierarchical model. You only need to override the entries where
the defaults are too vague or implausible for the task.

```{r priors}
pri <- prior(
  design = des,
  group_design = group_des,
  pmean = c(v = 1, v_lMd = 1, B = log(1), t0 = log(0.3), A = log(0.3)),
  psd = c(t0 = 0.3, A = 0.2, B = .5, sv_lMd = 0.3, t0_groupASD = 0.3),
  A = c(t0 = 0.1, A = 0.1)
)
```

There are three practical points to keep in mind here.

-   `pmean` and `psd` are overriding the default priors for selected
    coefficients. Everything not named here falls back to the package
    defaults.
-   `B`, `t0`, `A`, and `sv` are estimated on the log scale in the LBA
    implementation, so prior values for those parameters should be
    specified on the transformed scale.
-   The `A` argument in `prior()` is a prior hyperparameter for the
    hierarchical variance prior. It is not the same thing as fixing the
    model parameter `A`.

If you want to inspect the implied priors before fitting, a useful quick
check is:

```{r prior-check, eval=FALSE}
plot(pri, N = 1000, layout = c(2,2))
```

## Build the `emc` object

`make_emc()` combines the data, subject-level design, hierarchical group
design, and priors into the object that `fit()` needs.

```{r make-emc}
emc <- make_emc(data, design = des, prior_list = pri, group_design = group_des)
```

At this point the model is specified, but not yet estimated.

## Fit the model

The actual MCMC estimation step is:

```{r fit-model, eval=FALSE}
emc <- fit(emc, fileName = "samples/LBA.RData", cores_per_chain = 3)
```

This can take a while, so we load the saved fit.

```{r load-fit}
load("samples/LBA_manning.RData")
```

If you want to re-estimate the model yourself, simply evaluate the
`fit()` chunk and overwrite the saved file.

## Posterior predictive checking

Once the model is fit, the first question should be whether it
reproduces the main patterns in the observed data. `predict()` simulates
new datasets from the posterior, and `plot_density()` can overlay those
posterior predictive distributions on the real data.

```{r posterior-predictive, fig.height=6, fig.alt="Posterior predictive defective densities by condition and group"}
pp <- predict(emc, n_post = 50, n_cores = 9)

plot_density(
  emc,
  pp,
  factors = c("cond", "group"),
  functions = list(correct = acc_fun),
  defective_factor = "correct"
)
```

The posterior predictive distributions track the observed densities
well. The main patterns in accuracy and response-time shape are
reproduced in both groups and both conditions. The remaining mismatch is
small and is mainly visible in the far right tail, which is not unusual
in a model of this complexity.

## Summarize the posterior

For hierarchical regressions, `credint(emc, selection = "beta")` is a
good place to start because it returns credible intervals for the
group-level regression coefficients (including intercepts).

```{r posterior-intervals}
beta_ci <- credint(emc, selection = "beta")[[1]]

key_beta_ci <- beta_ci[
  c(
    "v", "v_groupASD",
    "v_lMd", "v_lMd_groupASD",
    "v_condeasy", "v_condeasy_groupASD",
    "v_lMd:condeasy", "v_lMd:condeasy_groupASD",
    "B", "B_groupASD",
    "t0", "t0_groupASD"
  ),
  ,
  drop = FALSE
]

knitr::kable(round(key_beta_ci, 3))
```

These intervals are easiest to read if you separate the *baseline terms*
from the *group-difference terms*.

-   Terms without `_groupASD` are the CTRL-group coefficients.
-   Terms with `_groupASD` tell you how much the ASD group shifts away
    from that CTRL baseline.

For this fitted model, the main instructional takeaways are:

-   `v_lMd` is strongly positive, so matching accumulators have much
    higher drift rates than mismatching accumulators. That is exactly
    what we would expect in a discrimination task.
-   `v_lMd:condeasy` is also positive, which means the match advantage
    is larger on easy trials.
-   `v_groupASD` is negative in this fit, suggesting lower baseline
    drift in the ASD group.
-   `B_groupASD` and `t0_groupASD` have intervals overlapping zero, so
    this model does not strongly separate the groups on threshold or
    non-decision time.

For `B` and `t0`, remember that the coefficients are on the transformed
scale used by the LBA implementation. In practice, the first pass is
usually about direction and uncertainty, not raw coefficient magnitude.

## Test specific effects with Bayes factors

Credible intervals are useful, but sometimes we want a direct test of a
point-null hypothesis such as whether a coefficient is equal to zero.
`hypothesis()` computes a Savage-Dickey Bayes factor for exactly that
question.

```{r bayes-factors}
targets <- c(
  "v_groupASD", "v_lMd_groupASD",
  "v_condeasy", "v_condeasy_groupASD",
  "v_lMd:condeasy", "v_lMd:condeasy_groupASD",
  "B_groupASD", "t0_groupASD"
)

bf_table <- data.frame(
  parameter = targets,
  BF10 = round(
    sapply(
      targets,
      function(tar) hypothesis(
        emc,
        selection = "beta",
        parameter = tar,
        do_plot = FALSE
      )
    ),
    3
  ),
  row.names = NULL
)

knitr::kable(bf_table)
```

This gives a summary of which effects have substantial evidence
against zero. In this model, the strongest evidence is for the
`v_lMd:condeasy` interaction, while the evidence for group differences
in `B` and `t0` is weak.

## Map coefficients back to interpretable cell summaries

Raw regression coefficients are useful for estimation, but they are not
always the easiest quantities to interpret. Often it is clearer to map
the posterior back to the condition means implied by the model.

The `map` argument in `credint()` does that bookkeeping for us. Here we
request:

-   drift rates for each combination of `group`, `cond`, and `lM`,
-   thresholds for each `group`,
-   non-decision times for each `group`,
-   and `sv` for matching and mismatching accumulators.

```{r mapped-posteriors}
mapped_ci <- credint(
  emc,
  map = list(v = c("group", "cond", "lM"), B = "group", t0 = "group", sv = "lM")
)[[1]]

knitr::kable(round(mapped_ci, 3))
```
